DBES Grand Challenge : High throughput, scalable solution for processing taxi trip data

### 2015-03-19 paper/experiments.tex
fixing some issues

Figure \ref{time_route} shows the execution times. For both setups execution time is higher for single instance due to lack of parallelism of the system. For distributed setup time get reduced with addition of new nodes and comes to a constant level. For single JVM set up highest throughput is achieved with two route processors. After that point time get slightly increases due to context switching.

@@

Figure \ref{time_profit} shows the execution times for both setups. In both setups, there is a higher latency at one instance due to lack of parallelism. Then for distributed set up execution times get reduced with addition of new resources and for single JVM setup it remains a constant.

@@

Figure \ref{latency_profit} shows the latencies. For single JVM, latencies slightly increased with the addition of new instances as expected. But for distributed setups there is a higher latencies compared to others. This latency is due to longer queues at the \textit{TopProfitProcessor}.

>>>>

Figure \ref{time_route} shows the execution times. For both setups execution time is higher for single instance due to lack of parallelism of the system. For distributed setup time get reduced with addition of new nodes and comes to a constant level. This means our solution scales up to 4 node level and maintains a constant throughput. This is due to two reasons. Firstly, \textit{RouteProcessor} is not a computing intensive process. Secondly, there is only one \textit{RouteEventEmitter} and the speed at which it can read and send messages becomes the bottleneck. For single JVM set up highest throughput is achieved with two route processors. After that point time get slightly increases due to context switching.

@@

Figure \ref{time_profit} shows the execution times for both setups. In both setups, there is a higher latency at one instance due to lack of parallelism. Then for distributed set up execution times get reduced with addition of new resources and for single JVM setup it remains a constant. This shows our solution scales up in a distributed setup. We can increase the system throughput by adding new nodes until \textit{ProfitEventEmitter} becomes the bottleneck.

@@

Figure \ref{latency_profit} shows the latencies. For single JVM, latencies slightly increased with the addition of new instances as expected. But for distributed setups there is a higher latencies compared to others. This latency is due to longer queues at the \textit{TopProfitProcessor}. In our message ordering process, we can only send a message to \textit{TopProfitProcessor} only if all queues are not empty. This allows some processes with less top ten profitable events to delay other process messages. This problem can  possibly fixed by adding a limit to the message queue size. On the other hand in a real time system, this event ordering many not required since event processing happens at real time.

### 2015-03-19 paper/paper.tex
fixing some issues

\title{Solving the grand challenge using an algorithmic and distributed approach}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\maketitle
\begin{abstract}
Processing complex queries on unbounded event streams in real time, is a challenge for many organizational data processing systems. These systems are expected to process data with minimal latency to generate real time events, and high through put to minimize the required hardware. In this regard DEBS Grand Challenge 2015 focus on evaluating two queries (frequent routes and  profitable cells) in real time with minimal latency and high throughput. These queries involve processing windows of thousands of records. Firstly, such processing demands efficient data structures and algorithms to minimize the processing over head. Secondly, system should partition data to process them in parallel to efficient use of hardware.

>>>>

\title{DBES Grand Challenge : High throughput, scalable solution for processing taxi trip data}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\maketitle
\begin{abstract}
Processing complex queries on unbounded event streams in real time, is a challenge for many organizational data processing systems. These systems are expected to process data with minimal latency to generate real time events, and high through put to minimize the required hardware. In this regard DEBS Grand Challenge 2015 focus on evaluating two queries (frequent routes and  profitable cells) in real time with minimal latency and high throughput. These queries involve processing windows of thousands of records. Firstly, such processing demands efficient data structures and algorithms to minimize the processing over head. Secondly, system should partition data to process them in parallel to make system scalable.

### 2015-03-19 paper/query_evaluation.tex
fixing some issues

\subsubsection{Expired values}
Route count of a particular route can be reduced due to expiration of some events. For an example, lets assume route count for route R1 is 10 and R1 is included in a top ten route event generated by a \textit{QueryEvaluator}. Then lets assume route count for R1 has been reduced to 7 (due to expiration events), when the next event is generated and hence R1 is not included in the next top ten route event. However since R1 value still is in the \textit{ResultAggregator} that may produce an event with this value. To avoid this problem, we send the removed route set details with the top ten route event generated by the \textit{QueryEvaluator}. In that case \textit{ResultAggregator} can first remove those routes.
\subsubsection{Out of order events}
\textit{ResultAggregator} may receive events in a different order than those were produced at the \textit{EventEmitter}. This out of order events may change the final output of the \textit{ResultAggregator}. We do message ordering at the \textit{ResultAggregator} to avoid this problem. In order to order messages, first we kept a message queue for each \textit{QueryEvaluator} and sent the message with least sequence number if a message is present in all queues.  However this resulted long queues due to speed difference of the different \textit{QueryEvaluator} processes. So we limited the queue size for 100. If a queue becomes 100, then we send all the messages less than the last message of that queue to \textit{ResultAggregator}.
\subsubsection{Intermediate top ten route events}
A route event increases the route count for that events' route. Further that can generate some expiration events which may reduce the route count for other routes. The top ten frequent routes can be changed due to these expiration events as well. Lets say in one \textit{QuerayEvaluator} case, a particular event cause two routes R1 and R2 to change and generates a top ten route change event. In the distributed setup these two routes can be in two \textit{QueryEvaluators} and that can cause two top ten route change events. These two events may result in two top ten route change events at \textit{ResultAggregator} as well instead of one. We could not fixed this issue, but we observe around 5\% of such events generated and system goes to the one \textit{QueryEveluator} result after all such event execution.

>>>>

\begin{enumerate}
	\item \textbf{Expired values} : Route count of a particular route can be reduced due to expiration of some events. For an example, lets assume route count for route R1 is 10 and R1 is included in a top ten route event generated by a \textit{QueryEvaluator}. Then lets assume route count for R1 has been reduced to 7 (due to expiration events), when the next event is generated and hence R1 is not included in the next top ten route event. However since R1 value still is in the \textit{ResultAggregator} that may produce an event with this value. To avoid this problem, we send the removed route set details with the top ten route event generated by the \textit{QueryEvaluator}. In that case \textit{ResultAggregator} can first remove those routes.
	\item \textbf{Out of order events} : \textit{ResultAggregator} may receive events in a different order than those are produced at the \textit{EventEmitter}. This out of order events may change the final output of the \textit{ResultAggregator}. We order messages at the \textit{ResultAggregator} to avoid this problem. In order to order messages, we keep a message queue for each \textit{QueryEvaluator} and sent the message with least sequence number to \textit{ResultAggregator},  if a message is present in all queues. 
	\item \textbf{Intermediate top ten route events} : A route event increases the route count for that events' route. Further that can generate some expiration events which may reduce the route count for other routes. The top ten frequent routes can be changed due to these expiration events as well. Lets say in one \textit{QuerayEvaluator} case, a particular event cause two routes R1 and R2 to change and generates a top ten route change event. In the distributed setup these two routes can be in two \textit{QueryEvaluators} and that can cause two top ten route change events. These two events may result in two top ten route change events at \textit{ResultAggregator} as well instead of one. Furether we observe only around 5\% of such events generated, and system goes to the one \textit{QueryEveluator} result after all such event are being executed. Solution for such a problem may depend on the nature in which a real application, uses these results in decision making process. Therefore we could not provide any fix for this problem.
\end{enumerate}

### 2015-03-19 paper/query_implementation.tex
fixing some issues

We have implemented profitable cells query in a similar manner to the frequent route query using  time windows. Complexity of the query raised following issues and we have solved them as given bellow.
\begin{enumerate}
	\item The profitability is calculated using the median profit and the number of empty taxis. In order to calculate median with O(log(n)) time complexity, we use two heaps (\textit{java.util.PriorityQueue}) . One is used to keep the values lesser than the median value (max heap) and other is to keep higher values (min heap). However the \texttt{delete} operation of \textit{java.util.PriorityQueue} takes O(n) time, since it traverse along the array to find the index of the object. We tried with our \textit{DynamicHeap} implementation to fix this. But that did not improved the performance. Firstly, these heap sizes does not grow over the value of 60 and there are same values repeating in the list increasing the probability of find a value without traversing the whole list. Secondly, our implementation create more objects compared to just keeping double values though the time complexity is O(log(n)). 
	\item In order to keep tract of empty taxis, we need to process both pickup cell and dropoff cell details for each message. One way of doing this is to send all events for same process instance and process both details with one event. However this avoid the possibility of partitioning data across different processes. We generate two events, one for pickup data and one for dropoff data to make the process parallel. 
	\item In order to handle expiration of dropoff events, we keep track of the taxi ids currently available in a particular cell. If a pickup event occurs before dropoff event expiration, then we remove that taxi id from that cell and reduce the empty taxi count. In such a case dropoff event expiration does not cause a reduction of taxi count.
\end{enumerate}

>>>>

We have implemented profitable cells query in a similar manner to the frequent route query using  time windows. Complexity of the query raised following issues and we have solved them as given bellow.
\begin{enumerate}
	\item The profitability is calculated using the median profit and the number of empty taxis. In order to calculate median with O(log(n)) time complexity, we use two heaps (\textit{java.util.PriorityQueue}) . One is used to keep the values lesser than the median value (max heap) and other is to keep higher values (min heap). However the \texttt{delete} operation of \textit{java.util.PriorityQueue} takes O(n) time, since it traverse along the array to find the index of the object. We tried with our \textit{DynamicHeap} implementation to fix this. But that did not improved the performance. Firstly, these heap sizes does not grow over the value of 145 and there are same values repeating in the list increasing the probability of find a value without traversing the whole list. Secondly, our implementation create more objects compared to just keeping double values though the time complexity is O(log(n)). 
	\item In order to keep tract of empty taxis, we need to process both pickup cell and dropoff cell details for each message. One way of doing this is to send all events for same process instance and process both details with one event. However this avoid the possibility of partitioning data across different processes. We generate two events, one for pickup data and one for dropoff data to make the process parallel. 
	\item In order to handle expiration of dropoff events, we keep track of the taxi ids currently available in a particular cell. If a pickup event occurs before dropoff event expiration, then we remove that taxi id from that cell and reduce the empty taxi count. In such a case dropoff event expiration does not cause a reduction of taxi count.
\end{enumerate}
